\documentclass[a4paper]{article}

\usepackage[margin=3.5cm]{geometry}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage[table,x11names]{xcolor}
\usepackage{logicproof}
\usepackage{url}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{xypic}

\usepackage{titlesec}
\usepackage{longtable}

% For restatables
\usepackage{thmtools}
\usepackage{thm-restate}
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}
\newtheorem*{remark}{Remark}
\declaretheorem[name=Exercise,numberwithin=section]{exc}

% Writing
\newcommand{\ie}{{i.e.}}
\newcommand{\eg}{{e.g.}}

\renewcommand\thepart{\Alph{part}}
\titleformat{\part}{\Large\filcenter\scshape}{\textnormal{\textbf{Part \thepart}}}{1em}{\\}

\title{\vspace{-2.5em}CO519 - Theory of Computing - Logic}
\author{Dominic Orchard \\
  {\small{School of Computing, University of Kent}}}

\date{Last updated on \today}

\begin{document}
\maketitle

\noindent
These are the accompanying notes for the logic part of CO519. They
provide a counterpart to the lectures, but they do not replace them; the
lectures will provide content, detail, and discussion not given in
these notes.  However, the notes also provide additional exercises and
examples, and some additional detail beyond what is assessed on this
course-- this will be pointed out when it happens.

These notes also act as a counterpart to the \textbf{course textbook},
\emph{Logic in Computer Science} (by Huth and Ryan) which I
recommend. We will only cover material from the first two chapters due
to the short length of this part of the course. I recommend reading
these chapters, but the assessable material for this course is
covered in lectures and these notes (excluding the appendices); the
textbook contains extra detail which is not assessed, but
worth learning.

If you spot any errors or have suggested edits, the notes are written
in LaTeX and are available on GitHub at
\url{http://github.com/dorchard/co519-logic}. Please fork and submit a
pull request with any suggested changes.

The \textbf{last page of these notes} provides a section for you to
write down any topics you do not yet understand, or feel like you need to
work on more. By writing here during the course, you can offload the task of remembering
what you need to work on. You can then look back at this list when you
are studying, and eventually cross of items as you get to grips with
the material.

\part{Propositional logic \& its natural deduction
      proofs}

\input{partA-notes.tex}

\part{Modelling systems using logic}
\setcounter{section}{0}

\input{partB-notes.tex}

\part{Satisfiability for propositional logic}
\setcounter{section}{0}

\input{partC-notes.tex}

\part{First-order logic and its natural deduction
  proofs}
\setcounter{section}{0}

\input{partD-notes.tex}

\newpage
\thispagestyle{empty}

\section*{Aide memoire}

\noindent
Write down any terms, symbols, concepts, processes, etc. that
you don't yet understand or would like to grasp better.
You can then cross these out as you make progress:
it might be 5 minutes after you first wrote the
item down here, or it might be 5 minutes before the exam.
This will help you to keep a record of the things you
need to work on. \\[0.5em]

\begin{center}
\hspace{-2em}\begin{tabular}{l|l}
  \textbf{Things I don't understand yet} &
   \textbf{Things I need to get better at} \\ \hline
  & \\
  & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
                                         & \\
             \end{tabular}
\end{center}


\end{document}